---
title: "Breast Cancer Classification"
author: "Roberto Ruiz Felix"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Issue
Given a data set of breast cancer imaging descriptions, are the tumors benign or malignant? 

## Import Libararies

```{r}
suppressPackageStartupMessages({
library(ggplot2)
library(dplyr)
library(plotly)
library(ggcorrplot)
library(caret)
library(tidyverse)
})
```

## Load Data
Mean, Standard Error, and 'worst' (largest) were computed for each image, resulting in 30 features. 
```{r}
data = read.csv("/Users/robertoruizfelix/Downloads/R Code/Breast Cancer Analysis/breast-cancer-wisconsin-data.csv")
head(data)
```

```{r}
str(data)
```

## Visualize all variables

```{r}
missing.val = colSums(is.na(data))
missing.val
```

```{r}
data.1 = data %>% as.data.frame() %>% 
  select_if(is.numeric) %>% gather(key = 'variable', value = 'value')

ggplot(data.1, aes(value)) + geom_density() + facet_wrap(~variable)
```

```{r}
ggplot(data, aes(x = diagnosis, fill = diagnosis)) +
  geom_bar() +
  labs(x = "Diagnosis", y = "Frequency", title = "Barplot of Diagnosis Column") + 
  scale_fill_manual(values = c("blue", "red"))
```
### Convert Diagnosis from 'B' & 'M' to 0 & 1
```{r}
data$diagnosis = ifelse(data$diagnosis == "M", 1, 0)
```

```{r}
suppressWarnings({
library(psych)})
pairs.panels(data, gap = 0, bg = c('red', 'blue', 'green')[data$diagnosis], pch = 21)
```

## Correlation Matrix
```{r}
ggcorrplot(cor(data[, 2:32]))
```

### Correlation Matrix of Mean data
```{r}
mean.data = data[, 2:12]
ggcorrplot(cor(mean.data))

suppressWarnings({
library(psych)})
pairs.panels(mean.data, gap = 0, bg = c('red', 'blue', 'green')[mean.data$diagnosis], pch = 21)
```


### Correlation Matrix of Standard Error data
```{r}
SE.data = data[, c(2, 13:22)]

ggcorrplot(cor(SE.data))

suppressWarnings({
library(psych)})
pairs.panels(SE.data, gap = 0, bg = c('red', 'blue', 'green')[SE.data$diagnosis], pch = 21)
```


### Correlation Matrix of Worst data
```{r}
worst.data = data[, c(2, 23:32)]
ggcorrplot(cor(worst.data))

suppressWarnings({
library(psych)})
pairs.panels(worst.data, gap = 0, bg = c('red', 'blue', 'green')[worst.data$diagnosis], pch = 21)
```


## Data Standardization & Matrix
```{r}
normalized.data = scale(data[, 3:32])
normalized.data.all = data.frame(data$diagnosis, normalized.data)
ggcorrplot(cor(normalized.data.all))
```
```{R}
pairs.panels(normalized.data.all, gap = 0, bg = c('red', 'blue', 'green')[normalized.data.all$data.diagnosis], pch = 21)
```
## PCA Analysis
```{r}
data.pca = prcomp(normalized.data.all, center = TRUE, scale = FALSE)
summary(data.pca)
pca.components = data.pca$x[, 1:3]
head(pca.components)
```

### PCA Analysis on scaling the diagnosis column 
```r
data.pca = prcomp(data[, 2:32], center = TRUE, scale = TRUE)
summary(data.pca)
```

## Scree Plot
```{r}
library(ggplot2)
variance.explained = (data.pca$sdev^2) / sum(data.pca$sdev^2)

scree.data = data.frame(Component = paste0("PC", 1:length(variance.explained)),
                          VarianceExplained = variance.explained)

scree.plot = ggplot(scree.data, aes(x = Component, y = VarianceExplained)) +
  geom_bar(stat = "identity", fill = "skyblue", width = 0.5) +
  labs(x = "Principal Component", y = "Proportion of Variance Explained",
       title = "Scree Plot") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(scree.plot) 
```

## KNN Model 
```{r}
library(caret)

# Convert diagnosis to factor (0/'B', 1/'M')
normalized.data.all$data.diagnosis = factor(normalized.data.all$data.diagnosis, levels = c("0", "1"))

# Step 1: Split data (80/20 split)
set.seed(123) # For reproducibility
train.index = createDataPartition(normalized.data.all$data.diagnosis, p = 0.8, list = FALSE)
train.data = normalized.data.all[train.index, ]
test.data = normalized.data.all[-train.index, ]

# Step 2: Extract the top 3 principal components
pca = prcomp(train.data[, -1], center = TRUE, scale. = FALSE)
pca.components = pca$x[, 1:3]

# Step 3: Tune the KNN model
train.control = trainControl(method = "cv", number = 10)

param.grid = expand.grid(k = seq(1, 100, by = 3))

# Train KNN model with different k values 
tuned.knn.model = train(
  x = pca.components,
  y = train.data$data.diagnosis,
  method = "knn",
  trControl = train.control,
  tuneGrid = param.grid
)
tuned.knn.model

# Step 4: Evaluate best model on testing set
test.pca = predict(pca, newdata = test.data[, -1])
test.pca.components = test.pca[, 1:3]
predictions = predict(tuned.knn.model, newdata = test.pca.components)

# View model performance
confusionMatrix(predictions, test.data$data.diagnosis)
```

## Model Visualization

```{r}
library(plotly)

combined.pca = rbind(pca.components, test.pca.components)
dataset = factor(c(rep("Training", nrow(pca.components)), rep("Test", nrow(test.pca.components))))
diagnosis = factor(c(train.data$data.diagnosis, test.data$data.diagnosis))
prediction = factor(predictions)

plot.data = data.frame(
  PC1 = combined.pca[, 1],
  PC2 = combined.pca[, 2],
  PC3 = combined.pca[, 3],
  Dataset = dataset,
  Diagnosis = diagnosis,
  Prediction = ifelse(dataset == "Training", NA, prediction)
)

fig = plot_ly(plot.data, x = ~PC1, y = ~PC2, z = ~PC3,
                color = ~Dataset,
                colors = c("blue", "red"),
                symbol = ~Diagnosis,
                symbols = c("circle", "square"),
                opacity = 0.7) %>%
  add_markers(
    text = ~paste("Dataset:", Dataset, "<br>",
                  "Diagnosis:", Diagnosis, "<br>",
                  "Prediction:", ifelse(is.na(Prediction), "N/A", as.character(Prediction)))
  ) %>%
  layout(
    scene = list(
      xaxis = list(title = "PC1"),
      yaxis = list(title = "PC2"),
      zaxis = list(title = "PC3")
    )
  )
fig
```


